---
title: Impact of Multiple Tetracycline Exposures on the Conjugative Transfer Efficiency of Antimicrobial Resistance Plasmids in Salmonella enterica
author: "Nneka Vivian Iduu"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
  md_document:
    variant: gfm
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Install and Load required libraries

install.packages("GEOquery", repos = "http://bioconductor.org/packages/3.14/bioc")
# Install and load required packages
#install.packages(c("ggplot2", "dplyr", "tidyr"))
#if (!requireNamespace("BiocManager", quietly = TRUE))
  #install.packages("BiocManager")
#BiocManager::install("apeglm")
#BiocManager::install("repR")
#install.packages("reshape2")
#install.packages("ez") 
#install.packages("writexl")
#install.packages("broom")
#install.packages("sjPlot")
#install.packages("ggplotify")
library(apeglm)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tidyverse)
library(pheatmap)
library(RColorBrewer) 
library(ggrepel)
library(DESeq2)
library(edgeR)
library(GEOquery)
library(limma)
library(umap)
library(psych)
library(readxl)
library(tibble)
library(reshape2)
library(ez)
library(lme4)
library(lmerTest)
library(irr)
library(rptR)
library(MCMCglmm)
library(MuMIn)
library(sjPlot)
library(VennDiagram)
library(grid)
library(rptR)
library(broom)
library(sjPlot)
library(webshot2)
library(writexl)
library(emmeans)
library(ggpubr)
library(ggplotify)

```
## Introduction

Antimicrobial resistance (AMR) in foodborne pathogens, particularly *Salmonella enterica*, 
represents a significant public health concern, with plasmid-mediated resistance transfer 
playing a major role in its dissemination. This study investigates the effects of 
tetracycline exposure on gene expression and conjugative plasmid transfer dynamics in a 
co-culture of *Salmonella enterica* and *Escherichia coli* over a 240-minute treatment period.

RNA-seq data were analysed using a time-course design across four time points 
(0, 30, 90, and 240 minutes) to:

- Identify differentially expressed genes between tetracycline-treated and control groups
- Assess repeatability of gene expression measurements across conditions and time points
- Characterise temporal regulation of tetracycline resistance and conjugative transfer genes

## Dataset

- **Source:** NCBI Gene Expression Omnibus (GEO)
- **Accession:** GSE261094
- **Organism:** *Salmonella enterica* 452 (donor) + *Escherichia coli* J53 (recipient)
- **Design:** Tetracycline treatment (8 µg/mL) vs Control (PBS)
- **Time points:** 0, 30, 90, 240 minutes post-exposure
- **Replicates:** 3 biological replicates per condition

## 1. Data Import and PreProcessing

```{r}

#import dataset..............
getwd()
Combined_count <- read_excel("Data/Combined_count.xlsx")


# Use column_to_rownames to set the first column as row names
Combined_count <- column_to_rownames(Combined_count, var = "...1")  # Assuming first column is named "gene_id"
view(Combined_count)
head(Combined_count) # Verify the structure
str(Combined_count)  # Check the structure of the object


#load sample informmation
sample_info <- read_excel("Data/sample_info.xlsx")
View(sample_info) 

# Use column_to_rownames to set the first column as row names
sample_info <- column_to_rownames(sample_info, var = "...1")  # Assuming first column is named "gene_id"
head(sample_info) # Verify the structure
```

## 2. Gene Expression Distribution and Visualization

```{r}
#create a deseq object and import the count data and sample info..............
dSet <- DGEList(counts = Combined_count, group = sample_info$Condition) 
# Check the structure of the DGEList
str(dSet)


# Apply filterByExpr.....................
keep <- filterByExpr(dSet, min.count = 5, min.total.count = 10, min.prop = 0.5)
dSet_sorted <- dSet[keep, ]  # Keep only the filtered genes

# Extract counts from the DGEList object
counts_matrix <- dSet_sorted$count
sum(counts_matrix == 0)

# Log-transform the counts (adding 1 to avoid log(0))
log_counts <- log2(counts_matrix + 1)
sum(log_counts == 0)
str(log_counts)  # Check the structure of the log-transformed counts
summary(log_counts)

# Extract  group information and group labels
group_info <- dSet_sorted$samples
group_labels <- group_info$group
unique(group_labels)

# Identify control and treatment columns based on sample_info
control_samples <- sample_info$Condition == "Control"  
treatment_samples <- sample_info$Condition == "Treatment"  

# Subset the log_counts matrix
control_log_counts <- log_counts[, control_samples]
treatment_log_counts <- log_counts[, treatment_samples]


# Check the distribution of data and save 
#png("Result/histogram_gene_expression.png", 
   # width = 10, height = 5, units = "in", res = 300)
par(mfrow = c(1, 2))

# Control Histogram
hist(control_log_counts, 
     main = "Control", 
     xlab = "log2 Gene Expression", 
     col = rgb(0, 0, 1, 0.5),  # Blue
     xlim = range(c(control_log_counts, treatment_log_counts), na.rm = TRUE), 
     breaks = 30,
     probability = TRUE)

# Treatment Histogram
hist(treatment_log_counts, 
     main = "Treatment", 
     xlab = "log2 Gene Expression", 
     col = rgb(1, 0, 0, 0.5),  # Red
     xlim = range(c(control_log_counts, treatment_log_counts), na.rm = TRUE), 
     breaks = 30,
     probability = TRUE)
#dev.off()  # Close and save the file

# Q-Q plot for log_counts
#png("Result/QQ_plots.png", 
    #width = 10, height = 5, units = "in", res = 300)
par(mfrow = c(1, 2))

qqnorm(control_log_counts, main = "Control")
qqline(control_log_counts, col = "blue")

qqnorm(treatment_log_counts, main = "Treatment")
qqline(treatment_log_counts, col = "red")

#dev.off()  # Close and save the file
```

## 3. Data Reshaping and Metadata Integration

```{r}
#Transpose data...........................
# Convert the log_counts matrix to a data frame
log_counts_df <- as.data.frame(log_counts)
summary(log_counts_df)

# Add the row names as a column called 'Gene'
long_combined <- log_counts_df %>%
  rownames_to_column(var = "Gene") %>%  # Convert row names to a column
  pivot_longer(cols = -Gene, names_to = "Sample", values_to = "Count")  # Pivot to long format

# Display the reshaped data
head(long_combined)

# Merge long_combined with sample_info
#long_combined <- long_combined %>%
  #left_join(sample_info, by = c("Sample" = "...1"))  # Adjust this as necessary based on your sample names

long_combined <- long_combined %>%
  left_join(sample_info %>% rownames_to_column(var = "Sample"), by = "Sample")

# Check the merged data
head(long_combined)
summary(long_combined)

# Ensure Condition and Time are factors
long_combined$Condition <- factor(long_combined$Condition)
long_combined$Time <- factor(long_combined$Time)
```

## 4. Linear Mixed Effects Model

```{r}
# Fit a linear mixed-effects model......................

#run for eithr time or conditon, and calculate the ICC intercept/total
mixed_model <- lmer(Count ~ Time * Condition + (1 | Gene), data = long_combined)

# Summarize the model
summary(mixed_model)

tab_model(mixed_model)

# First save as HTML
tab_model(mixed_model, file = "Result/mixed_model_summary.html")

# Then convert to PNG
webshot2::webshot("Result/mixed_model_summary.html", 
                  "Result/mixed_model_summary.png")
r.squaredGLMM(mixed_model)
```

### 4.1 Model Diagnostics and Predictions

```{r}
# Create a data frame for predictions
pred_data <- expand.grid(
  Time = factor(c(0, 30, 90, 240)),  # Convert Time to a factor
  Condition = factor(c("Control", "Treatment"))  # Convert Condition to a factor
)

# Calculate predicted values based on your mixed model
pred_data$Predicted <- predict(mixed_model, newdata = pred_data, re.form = NA)  # re.form = NA to ignore random effects

# View the predictions
print(pred_data)

# Create the plot with the group aesthetic
pred <- ggplot(pred_data, aes(x = Time, y = Predicted, color = Condition, group = Condition)) +
  geom_line() +  # Draw lines connecting the predicted values
  geom_point() +  # Add points for predicted values
  labs(title = "",
       x = "Time (minutes)",
       y = "Predicted Count") +
  theme_minimal()
plot(pred)

# Save to Results folder
# ggsave("Result/predicated_count_TvC.png",
   #    plot = pred,
    #   width = 7, height = 5, units = "in", dpi = 300)

#post-hoc comparison of fixed effect
emm <- emmeans(mixed_model, ~ Time * Condition)
summary(emm)
plot(emm)


# Extract residuals from the mixed model
residuals <- residuals(mixed_model)

# Plot residuals
plot(residuals)
abline(h = 0, col = "red")  # Add a horizontal line at zero

```

## 5. Repeatability Analysis

```{r repeatability-1, eval=FALSE, cache=TRUE}
repA <- rpt(Count ~ (1|Gene), grname = "Gene", data = long_combined,  datatype = "Gaussian", nboot = 1000, npermut = 0)
repA

# Save repeatability plot
# png("Result/repeatability_overall.png",
    width = 6, height = 5, units = "in", res = 300)
plot(repA)
#dev.off()


# Example for repeatability - Different approach
#rpt_result <- rptGaussian(Count ~ Condition * Time + (1 | Gene), 
   #                       grname = "Gene",
   #                       data = long_combined,
    #                      nboot = 1000, 
     #                     npermut = 100)
#print(rpt_result)

# Save repeatability plot 2
#png("Result/repeatability_overall2.png",
   # width = 6, height = 5, units = "in", res = 300)
#plot(rpt_result)
#dev.off()


# For control
# png("Result/repeatibility_control.png", width = 6, height = 5, units = "in", res = 300)
control_data <- subset(long_combined, Condition == "Control")
rep_control <- rpt(Count ~ (1|Gene), grname = "Gene", data = control_data, datatype = "Gaussian", nboot = 1000, npermut = 0 )
rep_control
#dev.off()

# For treatment
#png("Result/repeatibility_treatment.png", width = 6, height = 5, units = "in", res = 300)
treatment_data <- subset(long_combined, Condition == "Treatment")
rep_treatment <- rpt(Count ~ Time + (1|Gene), grname = "Gene", data = treatment_data, datatype = "Gaussian", nboot = 1000, npermut = 0)
rep_treatment
#dev.off()

# Display side by side using base R
par(mfrow = c(1, 2))
plot(rep_control, main = "Control")
plot(rep_treatment, main = "Treatment")
par(mfrow = c(1, 1))
```

### 5.1 Repeatability Time Point 0 Analysis

```{r repeatability-2, eval=FALSE, cache=TRUE}
#subset for Time 0
T_N_0 <- subset(long_combined, Time == 0)
head(T_N_0)
mixed_model1 <- lmer(Count ~ Condition + (1 | Gene), data = T_N_0)
tab_model(mixed_model1)
# Summarize the model
summary(mixed_model1)

# Save tab_model as HTML
tab_model(mixed_model1, 
          file = "Result/mixed_model_time0_summary.html")

# Save summary as text
capture.output(summary(mixed_model1), 
               file = "Result/mixed_model_time0_summary.txt")

r.squaredGLMM(mixed_model1)

repA_0 <- rpt(Count ~ (1|Gene), grname = "Gene", data = T_N_0,  datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_0


# For control condition
repA_con_0 <- rpt(Count ~ (1 | Gene), grname = "Gene", data = subset(T_N_0, Condition == "Control"), datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_con_0

# For treatment condition
repA_trt_0 <- rpt(Count ~ (1 | Gene), grname = "Gene", data = subset(T_N_0, Condition == "Treatment"), datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_trt_0

# Display side by side
par(mfrow = c(1, 3))
plot(repA_0, main = "Overall T0")
plot(repA_con_0, main = "Control T0")
plot(repA_trt_0, main = "Treatment T0")
par(mfrow = c(1, 1))
```

### 5.2 Repeatability Time Point 30 Analysis

```{r repeatability-3, eval=FALSE, cache=TRUE}
#subset for Time 30
T_N_30 <- subset(long_combined, Time == 30)
head(T_N_30)
mixed_model2 <- lmer(Count ~ Condition + (1 | Gene), data = T_N_30)
# Summarize the model
summary(mixed_model2)
tab_model(mixed_model2)

# Save tab_model as HTML
tab_model(mixed_model2, 
          file = "Result/mixed_model_time30_summary.html")

# Save summary as text
capture.output(summary(mixed_model2), 
               file = "Result/mixed_model_time30_summary.txt")

r.squaredGLMM(mixed_model2)


repA_30 <- rpt(Count ~ (1|Gene), grname = "Gene", data = T_N_30,  datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_30

# For control condition
repA_con_30 <- rpt(Count ~ (1 | Gene), grname = "Gene", data = subset(T_N_30, Condition == "Control"), datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_con_30

# For treatment condition
repA_trt_30 <- rpt(Count ~ (1 | Gene), grname = "Gene", data = subset(T_N_30, Condition == "Treatment"), datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_trt_30

#side by side display
par(mfrow = c(1, 3))
plot(repA_30, main = "Overall T30")
plot(repA_con_30, main = "Control T30")
plot(repA_trt_30, main = "Treatment T30")
par(mfrow = c(1, 1))
```

### 5.3 Repeatability Time Point 90 Analysis

```{r repeatability-4, eval=FALSE, cache=TRUE}
#subset for Time 90
T_N_90 <- subset(long_combined, Time == 90)
head(T_N_90)
mixed_model3 <- lmer(Count ~ Condition + (1 | Gene), data = T_N_90)
# Summarize the model
summary(mixed_model3)
# Save tab_model as HTML
tab_model(mixed_model3, 
          file = "Result/mixed_model_time90_summary.html")

# Save summary as text
capture.output(summary(mixed_model3), 
               file = "Result/mixed_model_time90_summary.txt")

r.squaredGLMM(mixed_model3)

# overall repeatability plot
repA_90 <- rpt(Count ~ (1|Gene), grname = "Gene", data = T_N_90, datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_90

# For control condition
repA_con_90 <- rpt(Count ~ (1 | Gene), grname = "Gene", data = subset(T_N_90, Condition == "Control"), datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_con_90

# For treatment condition
repA_trt_90 <- rpt(Count ~ (1 | Gene), grname = "Gene", data = subset(T_N_90, Condition == "Treatment"), datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_trt_90

# side by side display
par(mfrow = c(1, 3))
plot(repA_90, main = "Overall T90")
plot(repA_con_90, main = "Control T90")
plot(repA_trt_90, main = "Treatment T90")
par(mfrow = c(1, 1))

```

### 5.4 Repeatability Time Point 240 Analysis  

```{r repeatability-5, eval=FALSE, cache=TRUE}

#subset for Time 240
T_N_240 <- subset(long_combined, Time == 240)
head(T_N_240)
mixed_model4 <- lmer(Count ~ Condition + (1 | Gene), data = T_N_240)
# Summarize the model
summary(mixed_model4)
# Save tab_model as HTML
tab_model(mixed_model4, 
          file = "Result/mixed_model_time240_summary.html")

# Save summary as text
capture.output(summary(mixed_model4), 
               file = "Result/mixed_model_time240_summary.txt")

r.squaredGLMM(mixed_model4)

#overall repeatability
repA_240 <- rpt(Count ~ (1|Gene), grname = "Gene", data = T_N_240, datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_240

# For control condition
repA_con_240 <- rpt(Count ~ (1 | Gene), grname = "Gene", data = subset(T_N_240, Condition == "Control"), datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_con_240

# For treatment condition
repA_trt_240 <- rpt(Count ~ (1 | Gene), grname = "Gene", data = subset(T_N_240, Condition == "Treatment"), datatype = "Gaussian", nboot = 1000, npermut = 0)
repA_trt_240

# side by side display
par(mfrow = c(1, 3))
plot(repA_240, main = "Overall T240")
plot(repA_con_240, main = "Control T240")
plot(repA_trt_240, main = "Treatment T240")
par(mfrow = c(1, 1))

```

### 5.5 Repeatability Data Visualization

```{r}

# Repeatability data visualization
data <- data.frame(
  Time = rep(c("0", "30", "90", "240"), each = 2),
  Condition = rep(c("Control", "Treatment"), times = 4),
  Repeatability = c(0.898, 0.899, 0.794, 0.918, 0.920, 0.935, 0.907, 0.934),
  Error = c(0.003, 0.003, 0.005, 0.002, 0.002, 0.002, 0.002, 0.002)
)

# Load ggplot2 library
library(ggplot2)

# Create the barplot
repeatability_plot <-ggplot(data, aes(x = Time, y = Repeatability, fill = Condition)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7) +
  geom_errorbar(aes(ymin = Repeatability - Error, ymax = Repeatability + Error),
                position = position_dodge(0.7), width = 0.2) +
  scale_fill_manual(values = c("steelblue", "lightcoral")) +
  scale_y_continuous(limits = c(0, 1)) +  # Set y-axis range from 0 to 1
  theme_minimal() +
  labs(title = "",
       x = "Time-point (minutes)",
       y = "Repeatability",
       fill = "Condition") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    legend.title = element_text(size = 13)
  )
plot(repeatability_plot)
# Save to Results folder
#ggsave("Result/repeatability_barplot_TvC.png",
   #    plot = repeatability_plot,
    #   width = 7, height = 5, units = "in", dpi = 300)

# Ensure Time is a factor with proper levels
data$Time <- factor(data$Time, levels = c("0", "30", "90", "240"))

# Create the barplot
#repeatability_plot2 <- ggplot(data, aes(x = Time, y = Repeatability, fill = Condition)) +
 # geom_bar(stat = "identity", position = position_dodge(), width = 0.8) +
  #geom_errorbar(aes(ymin = Repeatability - Error, ymax = Repeatability + Error),
    #            position = position_dodge(0.7), width = 0.2) +
 # scale_fill_manual(values = c("steelblue", "lightcoral")) +
 # scale_y_continuous(limits = c(0, 1)) +
 # theme_minimal() +
 # labs(title = "",
   #    x = "Time Point",
    #   y = "Repeatability",
     #  fill = "Condition") +
 # theme(
  #  plot.title = element_text(hjust = 0.5, size = 14),
   # axis.text = element_text(size = 12),
  #  axis.title = element_text(size = 13),
   # legend.title = element_text(size = 13)
 # ) +
 # facet_wrap(~ Condition)  # Separate plots for each group (Treatment/Control)

# Save to Results folder
#ggsave("Result/repeatability_barplot_Timpoints.png",
  #     plot = repeatability_plot2,
    #   width = 7, height = 5, units = "in", dpi = 300)
```

## 6. Differential Expression Analysis

```{r,echo=TRUE}

# (1) Differential gene expression overall treatment and control of based on voom

# Create design matrix
design_main <- model.matrix(~ Condition + Time, data = sample_info)

# Estimate the dispersion
dSet_est <- estimateDisp(dSet_sorted, design_main)

# Apply voom transformation to stabilize variance
voom <- voom(dSet_est, design_main, plot = TRUE)

# Fit the linear model
fit_main <- lmFit(voom, design_main)

# Apply eBayes to compute moderated t-statistics
fit_main <- eBayes(fit_main)

# View the coefficients
colnames(fit_main$coefficients)

# Differential expression for overall control vs treatment
topTable_main <- topTable(fit_main, coef = "ConditionTreatment", number = Inf)

# Filter significant genes (p-value < 0.05 and logFC > 2)
sig_genes_main <- topTable_main[topTable_main$P.Value < 0.05 & abs(topTable_main$logFC) > 2, ]
```

### 6.1 Visualization of Diferentially Expressed Genes

```{r}
# Visualization of DEGS 
#Create a data frame for plotting
volcano_data <- topTable(fit_main, coef="ConditionTreatment", number=Inf)
volcano_data$Significance <- ifelse(volcano_data$P.Value < 0.05 & abs(volcano_data$logFC) > 2, "Significant", "Not Significant")

# Plot the volcano plot
DEGs <- ggplot(volcano_data, aes(x=logFC, y=-log10(P.Value), color=Significance)) +
  geom_point(alpha=1, size=1.5) +
  scale_color_manual(values=c("black", "red")) +
  labs(x="Log Fold Change", y="-Log10(p-value)", title="") +
  theme_minimal()

plot(DEGs)

# Save to Results folder
# ggsave("Result/volcano_plot_DEGs.png",
      # plot = DEGs,
     #  width = 7, height = 5, units = "in", dpi = 300)
```

### 6.2 Time-Course Differential Expression Analysis

```{r}
#DESIGN MATRIX for time points comparison between treatment and control interaction using voom

# Redefine the Time variable as a factor with explicit levels
sample_info$Time <- factor(sample_info$Time)

# Create the design matrix with separate columns for each time point and interaction terms
design <- model.matrix(~ Condition + Time + Condition:Time, data = sample_info)
print(design)


# Rename columns to replace ":" with "_"
colnames(design) <- gsub(":", "_", colnames(design))
print(colnames(design))

# Define contrasts based on desired comparison
contrasts <- makeContrasts(
  Treatment_vs_Control_Time0 = ConditionTreatment,
  Treatment_vs_Control_Time30 = ConditionTreatment_Time30,
  Treatment_vs_Control_Time90 = ConditionTreatment_Time90,
  Treatment_vs_Control_Time240 = ConditionTreatment_Time240,
  
  # Control Time points vs Time 0
  Control_Time30_vs_Time0 = Time30,
  Control_Time90_vs_Time0 = Time90,
  Control_Time240_vs_Time0 = Time240,
  
  levels = design
)


# Estimate dispersion
dSet_est <- estimateDisp(dSet_sorted, design)


# Fit the LM model
v <- voom(dSet_est, design, plot = TRUE)
fit <- lmFit(v, design)

# Apply the contrasts
fit <- contrasts.fit(fit, contrasts)
fit <- eBayes(fit)

colnames(fit$coefficients)


# Differential expression for each time point
topTable_0 <- topTable(fit, coef="Treatment_vs_Control_Time0", number=Inf)
sig_genes_0 <- topTable_0[topTable_0$P.Value < 0.05 & abs(topTable_0$logFC) > 2, ]
topTable_30 <- topTable(fit, coef="Treatment_vs_Control_Time30", number=Inf)
sig_genes_30 <- topTable_30[topTable_30$P.Value < 0.05 & abs(topTable_30$logFC) > 2, ]
topTable_90 <- topTable(fit, coef="Treatment_vs_Control_Time90", number=Inf)
sig_genes_90 <- topTable_90[topTable_90$P.Value < 0.05 & abs(topTable_90$logFC) > 2, ]
topTable_240 <- topTable(fit, coef="Treatment_vs_Control_Time240", number=Inf)
sig_genes_240 <- topTable_240[topTable_240$P.Value < 0.05 & abs(topTable_240$logFC) > 2, ]


# Differential expression in control group to time zero
Con_30v0 <- topTable(fit, coef="Control_Time30_vs_Time0", number=Inf)
sig_con_30v0 <- Con_30v0[Con_30v0$P.Value < 0.05 & abs(Con_30v0$logFC) > 2, ]
Con_90v0 <- topTable(fit, coef="Control_Time90_vs_Time0", number=Inf)
sig_con_90v0 <- Con_90v0[Con_90v0$P.Value < 0.05 & abs(Con_90v0$logFC) > 2, ]
Con_240v0 <- topTable(fit, coef="Control_Time240_vs_Time0", number=Inf)
sig_con_240v0 <- Con_240v0[Con_240v0$P.Value < 0.05 & abs(Con_240v0$logFC) > 2, ]
```

### 6.3 DEG Overlap between Treatment and Control

```{r}
#visualization
# Create a list of DEGs at each time point for Treatment and Control interaction

venn_data1 <- list(
  "TvC0" = rownames(sig_genes_0),
  "TvC30" = rownames(sig_genes_30),
  "TvC90" = rownames(sig_genes_90),
  "TvC240" = rownames(sig_genes_240)
)

venn.plot <- venn.diagram(
  venn_data1,
  filename = NULL,
  fill = c("red", "green", "blue", "purple"),
  alpha = 0.5,
  main = "Overlap of Differentially Expressed Genes of TvC"
)
grid.draw(venn.plot)

# Save to Results folder
#png("Result/venn_timepoints_TvC_DEGs.png",
  #  width = 7, height = 7, units = "in", res = 300)
#grid.draw(venn.plot)
#dev.off()
```

### 6.3.1 DEG Overlap of Control relative to baseline across timepoints

```{r}
#visualization
# Create a list of DEGs at each time point for Control interaction at different timepoints relative to baseline
venn_data2 <- list(
  "30 vs 0" = rownames(sig_con_30v0),
  "90 vs 0" = rownames(sig_con_90v0),
  "240 vs 0" = rownames(sig_con_240v0)
)

venn.plot2 <- venn.diagram(
  venn_data2,
  filename = NULL,
  fill = c("red", "green", "blue"),
  alpha = 0.5,
  main = "Overlap of Differentially Expressed Genes in Control"
)
grid.draw(venn.plot2)

# Save to Results folder
#png("Result/venn_timepoints_Control_DEGs.png",
   # width = 7, height = 7, units = "in", res = 300)
#grid.draw(venn.plot2)
#dev.off()
```

### 6.3.2 Treatment vs Control DEGs Overlap at Timepoint 30

```{r}

# Treatment vs Control DEGs compared to Control baseline at each time point

#T30 and C30
T30_C30 <- list(
  "T30" = rownames(sig_genes_30),
  "C30" = rownames(sig_con_30v0)
)

venn.plot.30 <- venn.diagram(
  T30_C30,
  filename = NULL,
  fill = c("red", "blue"),
  alpha = 0.5,
  main = "Overlap of DEGs between T30 and C30v0"
)
grid.draw(venn.plot.30)

#png("Result/venn_DEGs_T30_vs_C30v0.png",
  #  width = 6, height = 6, units = "in", res = 300)
#grid.draw(venn.plot.30)
#dev.off()
```

### 6.3.3 Treatment vs Control DEGs Overlap at Timepoint  90

```{r}
#T90 and C90
T90_C90 <- list(
  "T90" = rownames(sig_genes_90),
  "C90" = rownames(sig_con_90v0)
)

venn.plot.90 <- venn.diagram(
  T90_C90,
  filename = NULL,
  fill = c("red", "blue"),
  alpha = 0.5,
  main = "Overlap of DEGs between T90 and C90v0"
)
grid.draw(venn.plot.90)
#png("Result/venn_DEGs_T90_vs_C90v0.png",
  #  width = 6, height = 6, units = "in", res = 300)
#grid.draw(venn.plot.90)
#dev.off()


```

### 6.3.4 Treatment vs Control DEGs Overlap at Timepoint  240

```{r}
#T240 and C240
T240_C240 <- list(
  "T240" = rownames(sig_genes_240),
  "C240" = rownames(sig_con_240v0)
)

venn.plot.240 <- venn.diagram(
  T240_C240,
  filename = NULL,
  fill = c("red", "blue"),
  alpha = 0.5,
  main = "Overlap of DEGs between T240 and C240v0"
)
grid.draw(venn.plot.240)

#png("Result/venn_DEGs_T240_vs_C240v0.png",
  #  width = 6, height = 6, units = "in", res = 300)
#grid.draw(venn.plot.240)
#dev.off()
```

### 6.4 DEG Counts
### 6.4.1 DEG Counts in Control Group Relative to Baseline

```{r}
# Count of Upregulated and downregulated genes in control group comparison separate lists
upregulated <- list(
  "30 vs 0" = rownames(sig_con_30v0[sig_con_30v0$logFC > 2, ]),
  "90 vs 0" = rownames(sig_con_90v0[sig_con_90v0$logFC > 2, ]),
  "240 vs 0" = rownames(sig_con_240v0[sig_con_240v0$logFC > 2, ])
)
summary(upregulated)
downregulated <- list(
  "30 vs 0" = rownames(sig_con_30v0[sig_con_30v0$logFC < 2, ]),
  "90 vs 0" = rownames(sig_con_90v0[sig_con_90v0$logFC < 2, ]),
  "240 vs 0" = rownames(sig_con_240v0[sig_con_240v0$logFC < 2, ])
)
summary(downregulated)


# Count upregulated and downregulated genes for each comparison
up_counts <- sapply(upregulated, length)
down_counts <- sapply(downregulated, length)

# Combine the counts into a single data frame
gene_counts <- data.frame(
  TimePoint = rep(c("30 vs 0", "90 vs 0", "240 vs 0"), each = 2),
  Regulation = rep(c("Up", "Down"), times = 3),
  Count = c(up_counts, down_counts)
)

# View the gene counts
print(gene_counts)

# Ensure TimePoint is a factor and ordered correctly
gene_counts$TimePoint <- factor(gene_counts$TimePoint, levels = c("30 vs 0", "90 vs 0", "240 vs 0"))

# Plot the data as stacked bars
DEGs_Con <-ggplot(gene_counts, aes(x = TimePoint, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", position = "stack", width = 0.4) +   # Use "stack" for stacked bars
  labs(
    title = "DEGs of control against baseline",
    x = "Time Point",
    y = "Number of DEGs"
  ) +
  scale_fill_manual(values = c("Up" = "blue", "Down" = "red")) +
  theme_minimal()

plot(DEGs_Con)
# Save to Results folder
#ggsave("Result/DEGs_control_vs_baseline.png",
     #  plot = DEGs_Con,
      # width = 7, height = 5, units = "in", dpi = 300)
```

### 6.4.2 DEG Counts: Treatment vs Control at Each Time Point

```{r}
#Visualization of the number of differentially expressed genes between treatment and control at each timepoint

# Create the data frame
gene_counts <- data.frame(
  TimePoint = c("0 min", "30 min", "90 min", "240 min"),
  Upregulated = c(11, 6, 39, 21),
  Downregulated = c(6, 45, 83, 5)
)

# Reshape the data to long format for ggplot2
gene_counts_long <- reshape2::melt(
  gene_counts,
  id.vars = "TimePoint",
  variable.name = "Regulation",
  value.name = "Count"
)

# Ensure TimePoint is ordered correctly
gene_counts_long$TimePoint <- factor(gene_counts_long$TimePoint, levels = c("0 min", "30 min", "90 min", "240 min"))

# Adjust the order of levels for stacking
gene_counts_long$Regulation <- factor(gene_counts_long$Regulation, levels = c("Downregulated", "Upregulated"))

# Plot the data as a stacked bar plot
DEGs_TvC <-ggplot(gene_counts_long, aes(x = TimePoint, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", position = "stack", width = 0.6) +  # Adjust width for slimmer bars
  labs(
    title = "DEGs of treatment against control at each timepoint",
    x = "Time Point",
    y = "Number of DEGs"
  ) +
  scale_fill_manual(values = c("Downregulated" = "red", "Upregulated" = "blue")) +
  theme_minimal()

plot(DEGs_TvC)

# Save to Results folder
#ggsave("Result/DEGs of TvC Timepoint comparison .png",
      # plot = DEGs_TvC,
      # width = 7, height = 5, units = "in", dpi = 300)
```


### 6.5 Visualization of Gene Expression Patterns Across Time Points between Conditions

```{r}
# (Method 1) Combine all differential genes across timepoints

combined_df_tp <- rbind(
  mutate(topTable_0, Timepoint = 0),
  mutate(topTable_30, Timepoint = 30),
  mutate(topTable_90, Timepoint = 90),
  mutate(topTable_240, Timepoint = 240)
)

# Ensure the first column is treated as gene names by converting row names
combined_df_tp <- rownames_to_column(combined_df_tp, var = "Gene")

# Reshape data for heatmap
heatmap_df_data <- combined_df_tp %>%
  select(Gene, Timepoint, logFC) %>%
  pivot_wider(
    names_from = Timepoint, 
    values_from = logFC, 
    names_prefix = "Time_"
  )


# Convert to matrix, removing the Gene column
heatmap_df_matrix <- as.matrix(heatmap_df_data[, -1])


# Optional: Replace NA with 0 or mean
heatmap_df_matrix[is.na(heatmap_df_matrix)] <- 0


# Create the heatmap
pheatmap(heatmap_df_matrix, 
         cluster_rows = TRUE, 
         cluster_cols = FALSE,
         main = "")

# Save to Results folder
#png("Result/heatmap_all_DEGs.png",
   # width = 8, height = 10, units = "in", res = 300)
#pheatmap(heatmap_df_matrix,
        # cluster_rows = TRUE,
         #cluster_cols = FALSE,
       #  main = "")
#dev.off()

# (Method 2.) Combine all significant genes across timepoints
#combined_sig_tp <- rbind(
 # mutate(sig_genes_0, Timepoint = 0),
  #mutate(sig_genes_30, Timepoint = 30),
 # mutate(sig_genes_90, Timepoint = 90),
 # mutate(sig_genes_240, Timepoint = 240)
#)

# Ensure the first column is treated as gene names by converting row names
#combined_sig_tp <- rownames_to_column(combined_sig_tp, var = "Gene")

# Reshape data for heatmap
#heatmap_sig_data <- combined_sig_tp %>%
  #select(Gene, Timepoint, logFC) %>%
 # pivot_wider(
  #  names_from = Timepoint, 
   # values_from = logFC, 
   # names_prefix = "Time_"
  #)

# Convert to matrix, removing the Gene column
#heatmap_sig_matrix <- as.matrix(heatmap_sig_data[, -1])


# Remove rows with NA or Inf values
#heatmap_data_clean <- heatmap_sig_matrix %>%
 # filter(
  #  !if_any(starts_with("Time_"), ~is.na(.) | is.infinite(.))
 # )

# Optional: Replace NA with 0 or mean
# heatmap_sig_matrix[is.na(heatmap_sig_matrix)] <- 0

# Create the heatmap
#pheatmap(heatmap_sig_matrix, 
        # cluster_rows = TRUE, 
        # cluster_cols = TRUE,
        # main = "Gene Expression Changes Across Time Points")
```

### 6.6 Visualization of Significant Genes by Time Point
### 6.6.1 Visualization of Significant Genes by Time Point 0

```{r}
# visualization of significantly regulated genes at each time-points
# 1.load gene name for significant time_0
GSE261094_Sal452_RNA_Seq_T <- read_excel("Data/GSE261094_Sal452_RNA_Seq_T.xlsx", sheet = "sig_gene_o")


# Ensure the first column is treated as gene names by converting row names
sig_genes_0 <- rownames_to_column(sig_genes_0, var = "Gene_ID")


write_xlsx(sig_genes_0, "Result/sig_genes_0.xlsx")

# Merge data based on Gene ID
sig_genes_name_0 <- sig_genes_0 %>%
  inner_join(GSE261094_Sal452_RNA_Seq_T, by = "Gene_ID")

# Save
write_xlsx(sig_genes_name_0, "Result/sig_genes_name_0.xlsx")


# Add a column for direction (Up or Down regulation)
sig_genes_name_0  <- sig_genes_name_0 %>%
  mutate(Direction = ifelse(logFC > 2, "Upregulated", "Downregulated"))

# Create the plot
sig_0 <- ggplot(sig_genes_name_0, aes(x = logFC, y = reorder(Gene_name, logFC))) +
  geom_bar(stat = "identity", aes(fill = Direction), width = 0.7) +
  scale_fill_manual(values = c("Upregulated" = "blue", "Downregulated" = "red")) +
  labs(x = "Log Fold Change (logFC)", y = "Differenial Expressed Genes (time point 0)") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme(axis.text.y = element_text(size = 8))  # Adjust text size if necessary
plot(sig_0)

# Save to Results folder
#ggsave("Result/Significant Genes Timepoint 0.png",
      # plot = sig_0,
      # width = 7, height = 5, units = "in", dpi = 300)
```

### 6.6.2 Visualization of Significant Genes by Time Point 30

```{r}
# 2.load gene name for significant time_30
GSE261094_Sal452_RNA_Seq_T <- read_excel("Data/GSE261094_Sal452_RNA_Seq_T.xlsx", sheet = "sig_gene_30")



# Ensure the first column is treated as gene names by converting row names
sig_genes_30 <- rownames_to_column(sig_genes_30, var = "Gene_ID")
library(writexl)
write_xlsx(sig_genes_30, "Result/sig_genes_30.xlsx")

# Merge data based on Gene ID
sig_genes_name_30 <- sig_genes_30 %>%
  inner_join(GSE261094_Sal452_RNA_Seq_T, by = "Gene_ID")

# Save
write_xlsx(sig_genes_name_30, "Result/sig_genes_name_30.xlsx")


# Add a column for direction (Up or Down regulation)
sig_genes_name_30  <- sig_genes_name_30 %>%
  mutate(Direction = ifelse(logFC > 2, "Upregulated", "Downregulated"))

# Create the plot
sig_30 <- ggplot(sig_genes_name_30, aes(x = logFC, y = reorder(Gene_name, logFC))) +
  geom_bar(stat = "identity", aes(fill = Direction), width = 0.7) +
  scale_fill_manual(values = c("Upregulated" = "blue", "Downregulated" = "red")) +
  labs(x = "Log Fold Change (logFC)", y = "Differenial Expressed Genes (time point 30)") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme(axis.text.y = element_text(size = 8))  # Adjust text size if necessary

plot(sig_30)

# Save to Results folder
#ggsave("Result/Significant Genes Timepoint 30.png",
   #    plot = sig_30,
     #  width = 7, height = 5, units = "in", dpi = 300)
```

### 6.6.3 Visualization of Significant Genes by Time Point 90

```{r sig-genes-90, fig.width=12, fig.height=14}
# 3.load gene name for significant time_90
GSE261094_Sal452_RNA_Seq_T <- read_excel("Data/GSE261094_Sal452_RNA_Seq_T.xlsx", sheet = "sig_gene_90")

# Ensure the first column is treated as gene names by converting row names
sig_genes_90 <- rownames_to_column(sig_genes_90, var = "Gene_ID")

write_xlsx(sig_genes_90, "Result/sig_genes_90.xlsx")

# Merge data based on Gene ID
sig_genes_name_90 <- sig_genes_90 %>%
  inner_join(GSE261094_Sal452_RNA_Seq_T, by = "Gene_ID")

# Save
write_xlsx(sig_genes_name_90, "Result/sig_genes_name_90.xlsx")


# Add a column for direction (Up or Down regulation)
sig_genes_name_90  <- sig_genes_name_90 %>%
  mutate(Direction = ifelse(logFC > 2, "Upregulated", "Downregulated"))

# Create the plot
sig_90 <- ggplot(sig_genes_name_90, aes(x = logFC, y = reorder(Gene_name, logFC))) +
  geom_bar(stat = "identity", aes(fill = Direction), width = 0.7) +
  scale_fill_manual(values = c("Upregulated" = "blue", "Downregulated" = "red")) +
  labs(x = "Log Fold Change (logFC)", y = "Differenial Expressed Genes (time point 90)") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme(axis.text.y = element_text(size = 7))  # Adjust text size if necessary
plot(sig_90)

# Save to Results folder
# ggsave("Result/Significant Genes Timepoint 90.png",
     #  plot = sig_90,
      # width = 9, height = 10, units = "in", dpi = 300)
```

### 6.6.4 Visualization of Significant Genes by Time Point 240

```{r}
# 4.load gene name for significant time_240
GSE261094_Sal452_RNA_Seq_T <- read_excel("Data/GSE261094_Sal452_RNA_Seq_T.xlsx", sheet = "sig_gene_240")



# Ensure the first column is treated as gene names by converting row names
sig_genes_240 <- rownames_to_column(sig_genes_240, var = "Gene_ID")

write_xlsx(sig_genes_240, "Result/sig_genes_240.xlsx")

# Merge data based on Gene ID
sig_genes_name_240 <- sig_genes_240 %>%
  inner_join(GSE261094_Sal452_RNA_Seq_T, by = "Gene_ID")

# Save
write_xlsx(sig_genes_name_240, "Result/sig_genes_name_240.xlsx")


# Add a column for direction (Up or Down regulation)
sig_genes_name_240  <- sig_genes_name_240 %>%
  mutate(Direction = ifelse(logFC > 2, "Upregulated", "Downregulated"))

# Create the plot
sig_240 <- ggplot(sig_genes_name_240, aes(x = logFC, y = reorder(Gene_name, logFC))) +
  geom_bar(stat = "identity", aes(fill = Direction), width = 0.7) +
  scale_fill_manual(values = c("Upregulated" = "blue", "Downregulated" = "red")) +
  labs(x = "Log Fold Change (logFC)", y = "Differenial Expressed Genes (time point 240)") +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme(axis.text.y = element_text(size = 8))  # Adjust text size if necessary

plot(sig_240)

# Save to Results folder
# ggsave("Result/Significant Genes Timepoint 240.png",
     #  plot = sig_240,
     #  width = 7, height = 5, units = "in", dpi = 300)



```

```{r}
fs::dir_tree()
```

